Ion pairing controls rheological properties of ``processionary'' polyelectrolyte hydrogels

### 2016-10-25 PALSE/response.tex

\begin{quotationi}
Page 5, Fig. 1: Br end group of the polymer POH/PBr and Nu end group in PNu+Br- is missing in the structure.
\end{quotationi}

We thank the Reviewer for spotting this omission. Actually we are not completely sure that nucleophile substitution actually takes place on a tertiary carbon, but indeed a \ce{Nu+Br-} tail group is our best guess.

\begin{quotationi}
Page 5, chapter 2.1, beginning of 1st section (polymer analysis): Why is Mw mentioned there? I suggest that Mn is presented. Was Mn calculated via monomer conversion or end group analysis via 1H NMR? Mw/Mn = 1.089 should be rounded up to 1.09.
\end{quotationi}

This is indeed a difficult question ; although most papers dealing with polymer chemistry present the molecular weight calculated by NMR as a Mn, we personally think that Mw better fits: the methodology is indeed based on the integration of repeating unit protons, and heavy chains are overexpressed in this integration. However, on the basis that (1) The NMR calculated value can anyway, strictly speaking, neither be considered as a ``pure''  Mn or Mw (2) Mn agrees better with existing literature (3) Polydispersity is anyway low, we changed the sentence to ``Mn =\SI{8300}{\dalton} from NMR, \SI{5125}{\dalton} from GPC''.

@@

%we changed the text to give Mn instead of Mw

\begin{quotationi}
Page 5, chapter 2.1, end of 2nd section (peaks of Br-/I-, detection limits for F-/Cl-): Where do 89/91 and 121 come from? The peaks of Br- and I- should be 79/81 and 126. The content of the text does not fit to what is shown in S2 and S3.
\end{quotationi}
We thank the Reviewer for spotting this typo. We quite fortunately agree with the periodic table.

\begin{quotationi}
 It also should be found an alternative analysis method to proof the successful anion exchange for F- and Cl-.
\end{quotationi}
We think that the complete disappearance of any trace of Br- from the MS spectrum after addition of either NaCl or NaF is a clear evidence that the process occurred successfully. We initially thought of other alternative methods, namely atomic absorption  and elemental analysis: the former is no more available in our laboratories; and we know that the latter is practically really tricky on those highly hygroscopic polymers, as water tends to accumulate within the sample during either storage or preparation (as seen in the DSC spectrum) making it very difficult to get reliable data.

@@

\begin{quotationi}
Page 10, chapter 2.4.3: typing error (“the the correlation”)
\end{quotationi}

Fixed.

\begin{quotationi}
Supporting Information
Please insert 1H NMR spectrum of POH in order to understand the calculation of Mn.
\end{quotationi}

Thank you for the suggestion, the figure has been added as Figure S2; the caption depicts the methodology used for Dp/Mn calculation.

@@

First we would like to thank the reviewer for having carefully read our manuscript and provided valuable comments and suggestions.

\begin{quotationi}
The authors report on rheological characterization and theoretical interpretation of properties of polyelectrolyte assemblies prepared from molecules with a single anionic headgroup and cationic groups along the chain. The presented work including preparation of the polyelectrolytes, their rheological characterization in solution as well as theoretical interpretation, is considered to be conducted in a nice way, it indicate the role of counterions in affecting the assembly modes of the structures and novel assembly modes are suggested. There are, however, some concerns that should be addressed before publication can be recommended:

Although offering novel handles to generate molecular assemblies, and for this reason, is considered to be general interest, the suggestion for the particular structures being assembled could be further bolstered by scattering tools.
\end{quotationi}

In many instances we tried to perform DLS measurements on our samples and, indeed, signals were observed. However the obtained results were largely unreproducible (even for a given sample upon slight change of the concentration, not to mention when two different samples were used). An example is shown below.

\includegraphics[width=\textwidth]{DLS.png}
\textit{Figure: \textbf{a} DLS data and best fit \textbf{b} size distribution by diffusion intensity and \textbf{c} size distribution by number for PImBr (top row) and PPyrBr (bottom row) in water at 10\% weight concentrations}

As the Reviewer can see, dispersion was very high, and signal quality so low that cumulant fit analyses were absolutely irrelevant. In our opinion, there is nothing relevant in these data, this is why we finally chose not to include it in the SI. If the referee thinks that it might be worth adding some of these data to the article, we are ready to do it. However, we do know whether a reasonable interpretation could be provided from the data. We leave the referee judge; top is PImBr bottom PPyrBr. Although a pattern seems to exist, it seems difficult to us to get its exact meaning, and to separate the artefact from the fact\ldots

@@

\begin{quotationi}
In applying network theory to interpret the data, it appears that e.g. loose end corrections are not employed. What is the rational for that?
\end{quotationi}

To our understanding, loose end corrections describe a situation where pre-existing linear chains of size $M$ are randomly cross-linked with an average of $M_c$ monomers between cross links with $M_c\leq M/2$. The modulus then scales as
\begin{equation}
\frac{G}{k_\mathrm{B}T} = \frac{c}{M_c}  \left(1 - \frac{2M_c}{M}\right).
\end{equation}
For example if $M_c$ is only one third of $M$, then two third of the original linear chain are loose ends and do not contribute to the elasticity.

This is rather different from our situation where the number of monomers between cross links is larger or equal to the number of monomers in the original chain. This situation stems from the reactivity of the heads: a cross link necessarily occurs at the head of a polymer, preventing one of the possible loose end, but leaving the possibility of many loose ends at the tails of each polymer.

We can better compare our situation to a gel formed by branched polymers, for example bottle brush polymers. In this situation, Papagiannopoulos et al. (\href{http://doi.org/10.1021/bm060287d}{Biomacromolecules 2006} or \href{http://doi.org/10.1039/b714864j}{Faraday Discussions 2008}) have shown that the shear modulus scales as
\begin{equation}
\frac{G}{k_\mathrm{B}T} = \frac{c}{N_\mathrm{tot}}.
\end{equation}
where $N_\mathrm{tot}$ is the total number of monomers between cross-links, including side chains that are topological `loose ends'.

In other words, one can consider two possible definitions of ``cross link''. In Flory's definition, every branching point is a cross link, therefore a brush has many loose ends. In the more coarse-grained definition used by Papagiannopoulos et al., branching points leading to a loose end are not considered meaningful cross links, but every monomer of the side chains actually contribute to entropy and thus to elasticity.

We added a paragraph at the end of 2.4.2 to stress this difference.

\begin{quotationi}
In the introduction, the authors refer to interest in hydrogels in general is related to their application potential in biomedicine, but do not explicitly relate the particular conditions used to modulate properties also to such an application domain later in the manuscript. Are the various solution conditions studied compatible with such applications? Are the results sensitive to addition of a background electrolyte (e.g., similar to physiological ionic strength)?
\end{quotationi}

We thank the Reviewer for this excellent suggestion. We added a paragraph and Figure 8b to describe the predictions of our model concerning the consequences of a background electrolyte concentration similar to physiological ionic strength. We think it demonstrates quite well the power of our model as well as the robustness of our claims concerning potential biomedical applications. However making our gels biocompatible would demand a careful selection of the ion pairs to ensure lack of toxicity, a feature that is out of the scope of the present paper.

### 2016-10-27 ion_pairing_paper.tex
resubmitted version

\newcommand{\figSsec}{S1\xspace}
\newcommand{\figSmassPImBr}{S2\xspace}
\newcommand{\figSmassPPyBr}{S3\xspace}
\newcommand{\figSmassPImI}{S4\xspace}
\newcommand{\figSmassPPyI}{S5\xspace}
\newcommand{\figSfrequency}{S6\xspace}

@@

 & \noindent\large{Hassan Srour,\textit{$^{a}$} Martien Duvall Deffo Ayagou,\textit{$^{a}$} Thi Thanh-Tam Nguyen,\textit{$^{a}$}, Nicolas Taberlet,\textit{$^{b}$} S\'{e}bastien Manneville,\textit{$^{b}$} Chantal Andraud,\textit{$^{a}$} Cyrille Monnereau,\textit{$^{a}$} and Mathieu Leocmach$^{\ast}$\textit{$^{c}$}} \\%Author names go here instead of "Full name", etc.

@@

\footnotetext{\textit{$^{a}$~Univ Lyon, Ens de Lyon, Univ Claude Bernard, CNRS,
Laboratoire de Chimie, F-69342 Lyon, France. }}%Fax: XX XXXX XXXX; Tel: XX XXXX XXXX; E-mail: xxxx@aaa.bbb.ccc}}
\footnotetext{\textit{$^{b}$~Univ Lyon, Ens de Lyon, Univ Claude Bernard, CNRS,
Laboratoire de Physique, F-69342 Lyon, France. }}
\footnotetext{\textit{$^{c}$~Univ Lyon, Univ Claude Bernard, CNRS, Institut Lumi\`{e}re Mati\`{e}re, F-69622 Villeurbanne, France. E-mail: mathieu.leocmach@univ-lyon1.fr}}%Fax: XX XXXX XXXX; Tel: XX XXXX XXXX; E-mail: xxxx@aaa.bbb.ccc}}

%Please use \dag to cite the ESI in the main text of the article.
%If you article does not have ESI please remove the the \dag symbol from the title and the footnotetext below.
\footnotetext{\dag~Electronic Supplementary Information (ESI) available: GPC chromatogram of POH, HRMS spectra before and after metathesis, frequency dependent rheology. See DOI: 10.1039/b000000x/}
%additional addresses can be cited as above using the lower-case letters, c, d, e... If all authors are from the same address, no letter is required

@@

In the present report, we investigate the role of the counterion condensation on the mechanical properties of such hydrogels. In a first part we describe how we take advantage of our postfunctionalization approach to play systematically with the nature (aromatic or not) of the cationic repeating unit and the associated halide counterion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}) varied along the Hofmeister series. In the second and third parts, we show respectively the qualitative change in gel formability and the quantitative variations of mechanical properties of the aqueous dispersions obtained from these well-characterized polymers. In a fourth part, we rationalize this behaviour by proposing a microscopic model based on the idea that at low dissociation rates, cross-links (defined as a point where three or more polymers meet) are not separated by a single macromolecule but by several (up to hundreds) polymers in a supramolecular chain. We call this behaviour ``processionary'' in analogy to pine processionary caterpillar (\textit{Thaumetopoea pityocampa}) behaviour\cite{Fabre1916}. In particular, this model enables the quantification of the charge condensation rate from standard rheological measurements.%Finally, we discuss our results and possible applications of such smart hydrogels.

@@

We synthesise the phosphonate terminated polymer hereafter referred to as \ce{PBr} according to our previously reported methodology\cite{Srour2014,Appukuttan2012}. We obtain a well-controlled linear polymer ($M_w= \SI{8200}{\dalton}$ from NMR, \SI{5614}{\dalton} from GPC with $M_w/M_n = 1.08$; see Supplementary Materials and Supplementary Figure~\figSsec), with a degree of polymerisation $N_0=70$ (NMR). This polymer serves as a common scaffold from which we derive  all studied systems. Nucleophilic addition of N-methylimidazole or N-methylpyrrolidine to a heated solution of \ce{PBr} in THF affords the corresponding poly(imidazolium) and poly(pyrrolidinium) compounds. In these cases, as an inherent consequence of the structure of the starting material, charge neutrality is provided by bromide counter ions, and the polymer will be referred to as \ce{PIm+Br-} and \ce{PPyr+Br-}, respectively (Figure~\ref{sch:synthesis}).

We performed anionic metathesis by pouring an aqueous solution of \ce{PIm+Br-} or \ce{PPyr+Br-} into a saturated aqueous solution of the different sodium halides (\ce{NaF}, \ce{NaCl}, \ce{NaI}) (Figure~\ref{sch:metathesis}). In the cases of \ce{PIm+F-}, \ce{PIm+Cl-} , \ce{PPyr+F-} and \ce{PPyr+Cl-}, we obtain a turbid suspension immediately after addition. After extensive dialysis of the resulting mixture against deionized water and lyophilization we recover in high yields the different \ce{PIm+X-} and \ce{PPyr+X-} (where \ce{X}=\ce{F}, \ce{Cl}, \ce{I}). In order to bring evidence for the efficiency of the ionic metathesis, we submit the resulting materials to negative-mode high-resolution mass spectrometry (HRMS, see Supplementary Materials). We unambiguously assess complete displacement of bromide counterions by the full disappearance of the diisotopic mass peak (89/\SI{91}{\dalton}, Supplementary Figures~\figSmassPImBr and \figSmassPPyBr). We further confirm substitution by iodide by the concomitant apparition of a characteristic monoisotopic peak (\SI{121}{\dalton}, Supplementary Figures~\figSmassPImI and \figSmassPPyI), while fluoride and chloride anions signals, which are below the detection limits of the spectrometer, could not be observed in the corresponding polymers (not shown).

@@

\subsection{Water swelling, counterion condensation and solvent quality}
Water-swelling properties turn out to be extremely dependent on the polymer composition. With bromide or iodide conterions, both poly(pyrrolidinium) and poly(immidazolium) polymers afford homogeneous and optically transparent gels upon swelling with deionised water. We observe a limited swelling for \ce{PPyr+Cl-}, resulting in a granular, inhomogeneous gel. With fluoride counterions, or in the case of \ce{PIm+Cl-}, we observe neither swelling nor dissolution. We recover a 7.4~\%~wt suspension as a biphasic mixture of its individual components.

@@

\subsection{Rheological measurements}
\begin{figure*}
\includegraphics{strain_rsc.pdf}%
\caption{Storage modulus $G^\prime$ ($\bullet$) and loss modulus $G^{\prime\prime}$ ($\square$) measured through oscillatory shear experiments plotted against the strain amplitude $\gamma$. The moieties change with rows and the counterions with columns. The fixed frequency is $f=\SI{1}{\hertz}$. All samples are at 7.4~\%~wt. The gray area encompasses the plastic regime that lies between the linear regime at very low amplitudes and the fluid-like shear-thinning regime at high strain amplitude.}%
\label{fig:strain}%
\end{figure*}

@@

The mechanical properties at low strain evolve with the factors influencing counter-ion condensation. Imidazolium based gels are much weaker than their pyrrolidinium counterpart. Comparison between \ce{PIm+Br-} and \ce{PPyr+Br-} is particularly illustrative of this trend, as their $G^\prime$ values (\SI{13}{\pascal} and \SI{8100}{\pascal}, resp.) differ by almost three orders of magnitude. On the bottom row of Figure~\ref{fig:strain} we can compare the rheological responses of poly(pyrrolidinium)-based hydrogels for the three counterions allowing dissolution in water. Gels with bromide and iodide counterions show roughly the same modulus value at vanishing strain for both $G^\prime$ and $G^{\prime\prime}$. By contrast, the gel with chloride counterions is weaker by more than one order of magnitude. In other words, counterion condensation is correlated to softer gel. This trend is confirmed between \ce{PIm+Br-} and \ce{PIm+I-}, the former being an order of magnitude softer than the later.

@@

Let us note $N$ the number of monomers between two crosslinks or entanglements. If $N>g$ the ``procession'' of polymer chains performs a random walk of correlation blobs, each being a self-avoiding walk. If $g>N>g_\mathrm{scr}$ the procession performs only a self-avoiding walk. In any case, each procession is thus an entropic spring of constant stiffness $k_\mathrm{B}T$. The number density of procession is $c/N$. Therefore, the shear modulus at low strain is given by:
\begin{equation}
G = \frac{c}{N}k_\mathrm{B}T.
\label{eq:G}
\end{equation}
Because of the different weights of the pyrrolidinium and imidazolium moieties and of the different counterions, $c$ is not constant as we chose to conduct our experiments at constant weight fraction $w$ of polymer (generally used to quantify gelation ability of a given gelator). This explains why the respective modulus or critical strains of \ce{PPyr+I-} (heavier, lower number density) and \ce{PPyr+Br-} are in reverse order with respect to the Hofmeister series. Taking into account the molecular mass $M$ of each polymer, the Avogadro number $\mathcal{N_A}$ and  knowing the density $d$ of the solvent, we obtain the number $n = N/N_0$ of chains between cross-link
\begin{equation}
n = \frac{\mathcal{N_A}}{M} w d \frac{k_\mathrm{B}T}{G}.
\label{eq:n}
\end{equation}

@@

@@

As the material is strained, individual processions are stretched, starting from the larger scales~\cite{Pincus1976}. As sketched in Figure~\ref{fig:stretch}, when the the correlation blobs are fully stretched, the procession is a linear assembly of electrostatic blobs. Stretching the chain further means exposing more thermal blobs to the solvent, a process that cannot be expressed using the model of a spring of constant stiffness. This condition is thus the limit between the linear and non linear regimes.

For all our samples, the linear regime is narrow $\gamma_0\approx 10^{-2}\ll 1$ if existing. It means that the procession is just long enough to start performing the self-avoiding walk. Incidentally, this implies that the procession is too short to perform the random walk. In other words the number $N$ of monomers between crosslinks is comparable to $g_\mathrm{scr}$. Using equation (\ref{eq:gscr}), this condition translates into
\begin{equation}
\left(\frac{B}{B_0}\right)^3 = 1 + \frac{B}{B_s}
\label{eq:solveB}
\end{equation}
with $B_s = c/(u c_s g_\mathrm{K})$ and $B_0 = b c^{1/3} N^{2/3} g_\mathrm{K}^{-1} = (bc/g_\mathrm{K}) \left(k_\mathrm{B}T/G\right)^{2/3}$. Here, we do not have salt per se, however negatively charged phosphonate heads and their counterions play the same role by contributing to screening. We thus have two ``salt'' charges per chain. Thus $c_s/c = 2/N_0$ and $B_s = N_0/(2u g_\mathrm{K}) \approx 12$. $B_0$ is obtained from $G^\prime(\gamma\rightarrow 0)$ for each sample.

@@

\begin{figure}
\includegraphics{phase_diag.pdf}
\caption{Conformation of semi-dilute polyelectrolytes at rest at various scales as a function of counterion condensation parameter $A$. The phase diagram is given for a reduced temperature $\tau=0.46$ and for the monomer number density of \ce{PIm+Br-}, although $c$ is slightly different for each sample. Each vertical line represents a sample.}
\label{fig:phasediag}
\end{figure}

@@

Figure~\ref{fig:phasediag} summarizes the different conformations of a procession as a function of scale and counterion condensation for our experimental $c$ and $\tau$. The most striking feature of this phase diagram is the narrowing of the regime between $D$ and $r_\mathrm{scr}$ (white region in Figure~\ref{fig:phasediag}) as counterion condensation increases. As the size of the electrostatic blob crosses the screening length, electrostatic repulsion is no longer able to sustain solubility. This abrupt transition between an uniform gel and a precipitate is what separates our five soluble samples from insoluble \ce{PPyr+F-}, \ce{PIm+Cl-} and \ce{PIm+F-}. Our rheological measurements probe the structure of the procession from the Kuhn length to the distance between crosslinks that is approximately $r_\mathrm{scr}$ (upper limit of the white region in Figure~\ref{fig:phasediag}). \ce{PPyr+Cl-} and \ce{PIm+I-} display the typical structure of weakly charged polyelectrolytes in poor solvent. \ce{PPyr+I-} and \ce{PPyr+Br-} lie right in the crossover to strongly charged regime. Their correct description is possible only through Equation~(\ref{eq:geDfromB}). Finally \ce{PIm+Br-} is extremely close to the transition to insolubility (within experimental errors) which explains its extreme softness and wide plastic regime.

@@

>>>>

\newcommand{\figSsec}{S1\xspace}
\newcommand{\figSNMR}{S2\xspace}
\newcommand{\figSmassPImBr}{S3\xspace}
\newcommand{\figSmassPPyBr}{S4\xspace}
\newcommand{\figSmassPImI}{S5\xspace}
\newcommand{\figSmassPPyI}{S6\xspace}
\newcommand{\figSfrequency}{S7\xspace}

@@

 & \noindent\large{Hassan Srour,\textit{$^{a}$} Martien Duvall Deffo Ayagou,\textit{$^{a}$} Thi Thanh-Tam Nguyen,\textit{$^{a}$}, Nicolas Taberlet,\textit{$^{b}$} S\'{e}bastien Manneville,\textit{$^{b}$} Chantal Andraud,\textit{$^{a}$} Cyrille Monnereau,$^{\ast}$\textit{$^{a}$} and Mathieu Leocmach$^{\ast}$\textit{$^{c}$}} \\%Author names go here instead of "Full name", etc.

@@

\footnotetext{\textit{$^{a}$~Univ Lyon, Ens de Lyon, Universit\'e Claude Bernard Lyon 1, CNRS,
Laboratoire de Chimie, F-69342 Lyon, France. E-mail: cyrille.monnereau@ens-lyon.fr}}%Fax: XX XXXX XXXX; Tel: XX XXXX XXXX; E-mail: xxxx@aaa.bbb.ccc}}
\footnotetext{\textit{$^{b}$~Univ Lyon, Ens de Lyon, Universit\'e Claude Bernard Lyon 1, CNRS,
Laboratoire de Physique, F-69342 Lyon, France. }}
\footnotetext{\textit{$^{c}$~Univ Lyon, Universit\'e Claude Bernard Lyon 1, CNRS, Institut Lumi\`ere Mati\`ere, F-69622, VILLEURBANNE, France. E-mail: mathieu.leocmach@univ-lyon1.fr. Twitter: @LamSonLeoc}}%Fax: XX XXXX XXXX; Tel: XX XXXX XXXX; E-mail: xxxx@aaa.bbb.ccc}}

%Please use \dag to cite the ESI in the main text of the article.
%If you article does not have ESI please remove the the \dag symbol from the title and the footnotetext below.
\footnotetext{\dag~Electronic Supplementary Information (ESI) available: GPC chromatogram and 1H NMR of POH, HRMS spectra before and after metathesis, frequency-dependent rheology. See DOI: 10.1039/b000000x/}
%additional addresses can be cited as above using the lower-case letters, c, d, e... If all authors are from the same address, no letter is required

@@

In the present report, we investigate the role of the counterion condensation on the mechanical properties of such hydrogels. In a first part we describe how we take advantage of our postfunctionalization approach to play systematically with the nature (aromatic or not) of the cationic repeating unit and the associated halide counterion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}) varied along the Hofmeister series. In the second and third parts, we show respectively the qualitative change in gel formability and the quantitative variations of mechanical properties of the aqueous dispersions obtained from these well-characterized polymers. In a fourth part, we rationalize this behaviour by proposing a microscopic model based on the idea that at low dissociation rates, cross-links (defined as a point where three or more polymers meet) are not separated by a single macromolecule but by several (up to hundreds) polymers in a supramolecular chain. We call this behaviour ``processionary'' in analogy to pine processionary caterpillar (\textit{Thaumetopoea pityocampa}) behaviour\cite{Fabre1916}. In particular, this model enables the quantification of the charge condensation rate from standard rheological measurements. \textcolor{blue}{In a fifth part we apply this model to discuss the consequences of a background electrolyte at physiological ionic strength.} %Finally, we discuss our results and possible applications of such smart hydrogels.

@@

We synthesise the phosphonate terminated polymer hereafter referred to as \ce{PBr} according to our previously reported methodology\cite{Srour2014,Appukuttan2012}. We obtain a well-controlled linear polymer (\textcolor{blue}{$M_n= \SI{8300}{\dalton}$ from NMR, \SI{5125}{\dalton} from GPC with $M_w/M_n = 1.09$}; see Supplementary Materials and Supplementary Figures~\figSsec and~\figSNMR), with a degree of polymerisation $N_0=70$ (NMR). This polymer serves as a common scaffold from which we derive  all studied systems. Nucleophilic addition of N-methylimidazole or N-methylpyrrolidine to a heated solution of \ce{PBr} in THF affords the corresponding poly(imidazolium) and poly(pyrrolidinium) compounds. In these cases, as an inherent consequence of the structure of the starting material, charge neutrality is provided by bromide counter ions, and the polymer will be referred to as \ce{PIm+Br-} and \ce{PPyr+Br-}, respectively (Figure~\ref{sch:synthesis}).

We performed anionic metathesis by pouring an aqueous solution of \ce{PIm+Br-} or \ce{PPyr+Br-} into a saturated aqueous solution of the different sodium halides (\ce{NaF}, \ce{NaCl}, \ce{NaI}) (Figure~\ref{sch:metathesis}). In the cases of \ce{PIm+F-}, \ce{PIm+Cl-} , \ce{PPyr+F-} and \ce{PPyr+Cl-}, we obtain a turbid suspension immediately after addition. After extensive dialysis of the resulting mixture against deionized water and lyophilization we recover in high yields the different \ce{PIm+X-} and \ce{PPyr+X-} (where \ce{X}=\ce{F}, \ce{Cl}, \ce{I}). In order to bring evidence for the efficiency of the ionic metathesis, we submit the resulting materials to negative-mode high-resolution mass spectrometry (HRMS, see Supplementary Materials). We unambiguously assess complete displacement of bromide counterions by the full disappearance of the diisotopic mass peak (79/\SI{81}{\dalton}, Supplementary Figures~\figSmassPImBr and \figSmassPPyBr). We further confirm substitution by iodide by the concomitant apparition of a characteristic monoisotopic peak (\SI{126}{\dalton}, Supplementary Figures~\figSmassPImI and \figSmassPPyI), while fluoride and chloride anions signals, which are below the detection limits of the spectrometer, could not be observed in the corresponding polymers (not shown).

@@

\subsection{Water swelling, counterion condensation and solvent quality}
Water-swelling properties turn out to be extremely dependent on the polymer composition. With \ce{Br-} or \ce{I-} conterions, both poly(pyrrolidinium) and poly(immidazolium) polymers afford homogeneous and optically transparent gels upon swelling with deionised water. We observe a limited swelling for \ce{PPyr+Cl-}, resulting in a granular, inhomogeneous gel. With \ce{F-} counterions, or in the case of \ce{PIm+Cl-}, we observe neither swelling nor dissolution. We recover a 7.4~\%~wt suspension as a biphasic mixture of its individual components.

@@

\subsection{Rheological measurements}
\begin{figure*}
\includegraphics{strain_rsc.pdf}%
\caption{Storage modulus $G^\prime$ ($\bullet$) and loss modulus $G^{\prime\prime}$ ($\square$) measured through oscillatory shear experiments plotted against the strain amplitude $\gamma$. The moieties change with rows and the counterions with columns. The fixed frequency is $f=\SI{1}{\hertz}$. All samples are at 7.4~wt\%. The gray area encompasses the plastic regime that lies between the linear regime at very low amplitudes and the fluid-like shear-thinning regime at high strain amplitude.}%
\label{fig:strain}%
\end{figure*}

@@

\textcolor{blue}{We observe that} the mechanical properties at low strain evolve with the factors influencing counter-ion condensation. Imidazolium based gels are much weaker than their pyrrolidinium counterpart. Comparison between \ce{PIm+Br-} and \ce{PPyr+Br-} is particularly illustrative of this trend, as their $G^\prime$ values (\SI{13}{\pascal} and \SI{8100}{\pascal}, resp.) differ by almost three orders of magnitude. On the bottom row of Figure~\ref{fig:strain} we can compare the rheological responses of poly(pyrrolidinium)-based hydrogels for the three counterions allowing dissolution in water. Gels with bromide and iodide counterions show roughly the same modulus value at vanishing strain for both $G^\prime$ and $G^{\prime\prime}$. By contrast, the gel with chloride counterions is weaker by more than one order of magnitude. In other words, counterion condensation is correlated to softer gel. This trend is confirmed between \ce{PIm+Br-} and \ce{PIm+I-}, the former being an order of magnitude softer than the later.

@@

Let us note $N$ the number of monomers between two cross-links or entanglements. If $N>g$ the ``procession'' of polymer chains performs a random walk of correlation blobs, each being a self-avoiding walk. If $g>N>g_\mathrm{scr}$ the procession performs only a self-avoiding walk. In any case, each procession is thus an entropic spring of constant stiffness $k_\mathrm{B}T$. The number density of procession is $c/N$. Therefore, the shear modulus at low strain is given by:
\begin{equation}
G = \frac{c}{N}k_\mathrm{B}T.
\label{eq:G}
\end{equation}
Because of the different weights of the pyrrolidinium and imidazolium moieties and of the different counterions, $c$ is not constant as we chose to conduct our experiments at constant weight fraction $w$ of polymer (generally used to quantify gelation ability of a given gelator). This explains why the respective modulus or critical strains of \ce{PPyr+I-} (heavier, lower number density) and \ce{PPyr+Br-} are in reverse order with respect to the Hofmeister series. Taking into account the molecular mass $M$ of each polymer, the Avogadro number $\mathcal{N_A}$ and  knowing the density $d$ of the solvent, we obtain the number $n = N/N_0$ of chains between cross-link
\begin{equation}
n = \frac{\mathcal{N_A}}{M} w d \frac{k_\mathrm{B}T}{G}.
\label{eq:n}
\end{equation}

@@

\textcolor{blue}{We stress that $N$ in Equation~(\ref{eq:G}) is the total number of monomers between effective cross-links, including possible side chains and loose ends. In that we follow results from the literature of branched polymer gels~\cite{Papagiannopoulos2008} that are more tractable than the original theory from Flory~\cite{Flory1944} and Case~\cite{Case1960} for randomly cross-linked linear chains with loose end corrections, where every junction between two polymers is a cross-link.}

@@

As the material is strained, individual processions are stretched, starting from the larger scales~\cite{Pincus1976}. As sketched in Figure~\ref{fig:stretch}, when the correlation blobs are fully stretched, the procession is a linear assembly of electrostatic blobs. Stretching the chain further means exposing more thermal blobs to the solvent, a process that cannot be expressed using the model of a spring of constant stiffness. This condition is thus the limit between the linear and non linear regimes.

For all our samples, the linear regime is narrow $\gamma_0\approx 10^{-2}\ll 1$ if existing. It means that the procession is just long enough to start performing the self-avoiding walk. Incidentally, this implies that the procession is too short to perform the random walk. In other words the number $N$ of monomers between cross-links is comparable to $g_\mathrm{scr}$. Using Equation (\ref{eq:gscr}), this condition translates into
\begin{equation}
\left(\frac{B}{B_0}\right)^3 = 1 + \frac{B}{B_s}
\label{eq:solveB}
\end{equation}
with $B_s = c/(u c_s g_\mathrm{K})$ and $B_0 = b c^{1/3} N^{2/3} g_\mathrm{K}^{-1} = (bc/g_\mathrm{K}) \left(k_\mathrm{B}T/G\right)^{2/3}$. Here, we do not have salt per se, however negatively charged phosphonate heads and their counterions play the same role by contributing to screening. We thus have two ``salt'' charges per chain. Thus $c_s/c = 2/N_0$ and $B_s = N_0/(2u g_\mathrm{K}) \approx 12$. $B_0$ is obtained from $G^\prime(\gamma\rightarrow 0)$ for each sample.

@@

\begin{figure}
\includegraphics{phase_diag.pdf}
\caption{Conformation of semi-dilute polyelectrolytes at rest at various scales as a function of counterion condensation parameter $A$. Each vertical line represents a sample. The phase diagrams are given for a reduced temperature $\tau=0.46$. \textcolor{blue}{(a) In pure water. Background screening length $\kappa^{-1}$ comes from the \ce{PO3^2-} heads.  $c$ is set by \ce{PIm+Br-}. (b) With a background electrolyte at $c_s=\SI{100}{\milli M}$. $c$ is set by \ce{PPyr+Cl-}. Variations in $c$ values from sample to sample have little impact on the phase boundaries, except near the crossing of $r_\mathrm{scr}$ and $D$.}}
\label{fig:phasediag}
\end{figure}

@@

Figure~\ref{fig:phasediag}a summarizes the different conformations of a procession as a function of scale and counterion condensation for our experimental $c$ and $\tau$. The most striking feature of this phase diagram is the narrowing of the regime between $D$ and $r_\mathrm{scr}$ (white region in Figure~\ref{fig:phasediag}a) as counterion condensation increases. As the size of the electrostatic blob crosses the screening length, electrostatic repulsion is no longer able to sustain solubility. This abrupt transition between an uniform gel and a precipitate is what separates our five soluble samples from insoluble \ce{PPyr+F-}, \ce{PIm+Cl-} and \ce{PIm+F-}. Our rheological measurements probe the structure of the procession from the Kuhn length to the distance between cross-links that is approximately $r_\mathrm{scr}$ (upper limit of the white region in Figure~\ref{fig:phasediag}a). \ce{PPyr+Cl-} and \ce{PIm+I-} display the typical structure of weakly charged polyelectrolytes in poor solvent. \ce{PPyr+I-} and \ce{PPyr+Br-} lie right in the crossover to strongly charged regime. Their correct description is possible only through Equation~(\ref{eq:geDfromB}). Finally \ce{PIm+Br-} is extremely close to the transition to insolubility (within experimental errors) which explains its extreme softness and wide plastic regime.

@@

{\color{blue}
\subsection{Application to physiological ionic strength}

To demonstrate the versatility of our model we expose briefly the consequences of having a background electrolyte with ionic strength similar to the physiological one and with the same counterion (e.g. \ce{Na+X-}). We set the salt concentration $c_s$ to \SI{100}{\milli M}, implying a background Debye length $\kappa^{-1}\approx 13 b$ and $B_s$ close to unity. We show the resulting phase diagram in Figure~\ref{fig:phasediag}b. The background electrolyte lowers the horizontal asymptote of $r_\mathrm{scr}$, which strongly reduces the white region of the phase diagram. \ce{PIm+I-} and \ce{PIm+Br-} have now lost the possibility to form linear chains, such that their electrostatic blobs are now space-filling. Therefore these two gels should precipitate at biological ionic strength. By contrast \ce{PPyr+Br-} and \ce{PPyr+I-} would still have linear chains between $D$ and $r_\mathrm{scr}$ and would stay soluble in a robust gel state.

Let us consider the case of \ce{PPyr+Cl-} whose chloride counterions are interesting from a physiological point of view. If cross-links still occur at the scale of the now shorter screening length, we expect less monomers between cross-links and thus larger shear modulus. Combining Equation (\ref{eq:gscr}) and (\ref{eq:G}) we obtain a modulus of \SI{780}{\pascal}, almost four times larger than \ce{PPyr+Cl-} in pure water. However \ce{PPyr+Cl-} would be a limit case, stable for an ionic strength of \SI{100}{\milli M} but precipitating at \SI{150}{\milli M} which is still in the range of physiological ionic strength. To ensure gel stability, it would thus be necessary to use cationic moeties with even less counterion condensation than pyrrolidinium.
}
